---
title: "Italian Restaurant analysis"
output:
  pdf_document: default
  html_document: default
  word_document: default
date: "2025-01-07"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## My analysis starts here

* Dataset consists of 10000 observations and 9 variables for the calender year 2024.  
* Variables : Date ,Time ,Menu item , Category , Quantity purchased ,Price per item , Revenue , Payment method , Customer type
* No null observations
* Currency : Euro

```{r cars, message=FALSE, warning=FALSE,echo=T}

library(readr)
library(tidyverse)
library(readxl)
library(sqldf)
library(stringr)
library(knitr)
library(rpart)

italian_data <-read_excel('C:\\Users\\karab\\Videos\\JOB_Projects\\Italian_resteurant_dataset.xlsx')
head(italian_data)

summary(italian_data)

if( is.null(italian_data)){
  print('The data contains missing observation')
}else{"The data has no missing observations"}

```


```{r aee, message=FALSE, warning=FALSE, echo=T}
# Problem with date, convert to date
data<-italian_data
data$Date<-as.Date(data$Date)

data <-data %>%mutate(AM_PM=str_extract(Time,'AM|PM'),Month=months.Date(Date),
                       Week_day=weekdays(Date))%>%select(-c(Date,Time))


mat <-matrix(0,nrow = 3,ncol = 2)
mat[,1]<-c('Price','Revenue','Quantity purchased')
mat[,2]<-c(round(mean(data$`Price (per item)`),2) , round(mean(data$Revenue),2) , round(mean(data$Quantity)))
colnames(mat)<- c('-','Average')
kable(mat)
####
print('Total revenue for each Category')

revenue_by_category <- data %>% select(Category,`Price (per item)`,Revenue,Quantity) %>%
  group_by(Category) %>% summarise(Average_price=round(mean(`Price (per item)`),2) , Total_customers= length(Category),
  Quantities_purchased=sum(Quantity) ,Total_revenue=sum(Revenue)) %>% arrange(desc(Total_revenue))
kable(revenue_by_category)

revenue_by_category %>% ggplot(aes(x=Category,y=Total_revenue)) + geom_bar(stat='Identity',
aes(fill = Category)) + scale_y_continuous(labels =scales::comma ) + 
  geom_text(aes(label = Total_revenue),position = position_stack(vjust=0.9)) + 
  labs(title = 'Total revenue for each category', y=' Total revenue(Euro)') + 
  theme(plot.title = element_text(hjust = 0.5))+ theme_bw()

####
```


**Total revenue per Category for each menu item'**

```{r a, message=FALSE, warning=FALSE,echo=T}

Revenue_per_category_by_menu_item <- data %>% select(Category,Quantity,`Menu Item`,Revenue)%>%
  group_by(Category,`Menu Item`) %>% summarise(Total_customers=length(`Menu Item`),                    Total_quantity=sum(Quantity),Total_Revenue=round(sum(Revenue)))

kable(Revenue_per_category_by_menu_item)

Revenue_per_category_by_menu_item %>% ggplot(aes(x=`Menu Item`,y=Total_Revenue,fill = `Menu Item`)) +
  geom_bar(stat='Identity') +labs(title = 'Total revenue for each menu item',y= 'Total revenue (Euro)') + geom_text(aes(label=Total_Revenue),angle=50,position = position_stack(vjust=0.5)) +
  theme_dark()+theme(axis.text.x = element_text(angle = 45,hjust=1),
                     plot.title = element_text(hjust = 0.5)) + theme_classic()

####
```


**Customer time preference (AM or PM)**
```{r b,echo=T, message=FALSE, warning=FALSE}
customer_preferred_time <- data%>% select(AM_PM,Revenue) %>%
  group_by(AM_PM)%>%
summarise(Total_customers=length(AM_PM),Revenue=round(sum(Revenue)))

kable(customer_preferred_time)

customer_preferred_time %>% ggplot(aes(x=AM_PM,y=Revenue,fill = AM_PM)) +
  geom_bar(stat='Identity') + labs(title = ' Revenue by meridiem(time) preferrence ', 
  y='Total revenue (Euro)',x='Meridiem') + geom_text(aes(label = Revenue),
 position = position_stack(vjust=0.9)) + scale_y_continuous(labels = scales::comma) +theme_minimal()


#### 
```


**Customer Payment Preference**
```{r c,echo=T, message=FALSE, warning=FALSE}
payment_preference<- data%>% select(`Payment Method`) %>%group_by(`Payment Method`) %>%
  summarise(Total_customers=length(`Payment Method`)) %>% arrange(Total_customers)
kable(payment_preference)

payment_preference %>% ggplot(aes(x=`Payment Method`,y=Total_customers))+
  geom_point(col=c('red','black','blue'),size=4,shape=15)+
  labs( title = "Customer's payment method preference") +
  theme(plot.title = element_text(hjust=0.4))+theme_classic() +
  geom_text(aes(label = Total_customers),
            hjust=-0.5,col=c('red','black','blue'))+theme_gray() + theme_bw()
####
```


**customer type in terms of food distribution**
```{r d,echo=T, message=FALSE, warning=FALSE}
customer_type<-data %>% select(`Customer Type`)%>%group_by(`Customer Type`) %>%
  summarise(Total_customers=length(`Customer Type`)) %>% 
  arrange(Total_customers) %>%rename(Service_type=`Customer Type`)

kable(customer_type)

customer_type%>%ggplot(aes(x=Service_type,y=Total_customers)) +
  geom_point(col=c('red','black','purple'),size=5,shape=20) +theme_classic() + 
  geom_text(aes(label = Total_customers),hjust=-0.4,col=c('red','black','purple'))+
  labs(title = "Service type preference")+theme(element_text(hjust = 0.5))

#### 
```


**Monthly Revenue**
```{r e,echo=T, message=FALSE, warning=FALSE}
monthly_revenue <- data %>% select(Month,Revenue,Category) %>% group_by(Month,Category)%>%
  summarise(Total_revenue= round(sum(Revenue))) %>% 
  arrange(factor(Month,levels = c("January","February","March","April", "May" ,"June",
   "July", "August","September" ,"October","November","December" )))

kable(monthly_revenue%>%select(Month,Total_revenue) %>% group_by(Month) %>%
        summarise(Total_revenue=sum(Total_revenue)))

monthly_revenue %>% ggplot(aes(x=Month,y=Total_revenue,fill = Category))+
  geom_bar(stat = 'Identity',position = 'dodge') + labs(y='Total revenue (Euro)',
 title = 'Monthly revenue stacked by Category')+ 
  scale_fill_manual(values = c('Violet','#4B4','lightblue')) +
  theme_classic() +theme(axis.text.x=element_text(angle = 45,hjust = 1),
                         plot.title = element_text(hjust=0.5)) 

####
```


**Weekdays revenue**
```{r f,echo=T, message=FALSE, warning=FALSE}
weekly_revenue <- data %>% select(Week_day,Revenue) %>% group_by(Week_day) %>%
  summarise(Total_revenue= round(sum(Revenue))) %>% arrange(factor(Week_day,
  levels = c("Monday" ,"Tuesday","Wednesday" ,"Thursday","Friday","Saturday","Sunday" )))

kable(weekly_revenue)

weekly_revenue %>% ggplot(aes(x=Week_day,y=Total_revenue,fill = Week_day))+ 
  geom_bar(stat = 'Identity') + theme_minimal() +labs(title = 'Daily revenue trend',
  y='Total revenue (Euro)')+ theme(plot.title = element_text(hjust = 0.5),
 axis.text.x=element_text(angle = 45,hjust = 1)) +
  geom_text(aes(label = Total_revenue),angle=60,position = position_stack(vjust=0.7))

####

```


```{r kb,message=FALSE, warning=FALSE}

# Create classification target
average_revenue <- mean( data$Revenue)
data$High_Revenue <- as.factor( ifelse(data$Revenue > average_revenue, "Yes", "No"))

# Train decision tree
model <- rpart(High_Revenue ~ Category + Quantity + `Customer Type` + AM_PM,
               data = data, method = "class")

# Visualize the tree
plot(model)
text(model, use.n = TRUE)

```

### Insights

 
 * Food items from the *Main course* category generates more revenue(139581.5 Euros) than the *Appetizer* and *Dessert* categories, with Dessert generating the least revenue(50541 Euros) among the three categories
 
 
 * *Fettuccine Alfredo* and *Toramisu* are the most liked food items by customers, *Caprese Salad* is the least favourite
 
 
 * *Fettuccine Alfredo* ,*Spaghetti Carbonara* and *Lasagna* are the top three revenue generating food items, *Gelato* and *Panna Cotta* generate way less revenue
 
 
 * Most customers buy more in the Afternoon or Evening (PM) than they do in the Morning (AM), hence the Restaurant generates more revenue in post meridiem (PM)
 
 * Most customer prefer paying *cash* compared to the likes of *Credit card* and *Mobile payment*
 
 * Most customers prefer *Takeaways* than *Delivery* and *eating in-store*
 
 * *August* and *October* generated the most revenue with *June* and *February* generating lowest revenue
 
 * The revenue of the weekdays is quite negligible, the difference is barely visible, but *Sunday* seems to generate more revenue than the other weekdays, with *Monday*
 generating the least  revenue
 
 * Main courses are key for achieving high revenue — especially when customers order in larger quantities.
 
 * Appetizers and desserts almost never lead to high revenue on their own 
 
 * Promotions that encourage bulk orders of main courses (e.g., family meals, combos) could significantly boost revenue.




